### 2020-01-18 nlp_models.tex
updated NLP section

\paragraph{The Heavy Tailed Power Law Exponents in GPT and GPT2}

are very differenmt, with GPT2 having both a notably smaller mean $\alpha$, and far fewer, unusually large outliers.
Figure \ref{fig:gpt-alphs-hist} shows the empirical density (histogram) of $\alpha$
for all layers in GPT (blue) and GPT2 (red).  \charles{discuss more}

\begin{figure}
    \centering
    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha.hist.png}
       }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm.hist.png}
       }

@@

\charles{We have redo the spectral norm and remove the outliers}

\paragraph{The Correlation Flow in GPT and GPT2} also differs significantly between GPT and GPT2.
Figure \ref{fig:gpt-alpha-layer} plots $\alpha$ vs the layer id for each model.

@@

    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=4.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=4.5cm]{img/GPT-snorm-depth.png}
        \label{fig:resnet-snorm-layer}
    }
    \qquad
    \subfigure[log Alpha-Norm $\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$]{
        \includegraphics[width=4.5cm]{img/GPT-pnorm-depth.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

Notice the first embedding layer(s) have unusually large Spectral Norms,
reflecting a lack of normalization.  For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we layer all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier); including them does
not significantly alter the results here.

>>>>

\paragraph{Empirical Quality Metrics for GPT and GPT2}
We first compare the distribution of  Heavy Tailed Power Law exponents $\alpha$ in GPT and GPT2.
They are very differenmt, with GPT2 having both a notably smaller mean $\alpha$, and far fewer, unusually large outliers.
Figure \ref{fig:gpt-alphs-hist} shows the empirical density (histogram) of $\alpha$
for all layers in GPT (blue) and GPT2 (red).  \charles{discuss more}

\begin{figure}
    \centering
    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }

@@

Indeed, the $\alpha$ metric makes a good quality metric for comparing these models.
Figure \ref{fig:GPT-alpha-hist} compares the PL exponent $\alpha$ for the GPT and GPT2 models for all layers,
As expected, the improved GPT2 model has, on average, smaller $\alpha$ than the older GPT, with all
$\alpha\le6$.  Indeed, the deficient GPT model has numerous weight matrices with unsually large fitted exponents,
indicating that they are not Heavy Tailed at all.  Indeed, we may expect that a poorly trained model
will not exhibit consistent Heavy Tailed behavior in all layers.

However, as seen in Figure \ref{fig:GPT-snorm-hist},
 the poorly trained GPT model has many smaller (log) Spectral Norms $\log\Vert\mathbf{W}\Vert_{\infty}$
than the GPT2, which would be inconsistent with a belief that smaller Spectral Norm is always better.
Indeed, because there are so many anamously small $\Vert\mathbf{W}\Vert_{\infty}$,
it appears that the GPT model may be exhibiting a kind of rank collapse.
This is an extremely important observation because it demonstrates that while the Spectral Norm
may correlate well with predicted test error, it is not a good indicator of the overall quality of a model,
and using it as an empirical metric may give spurious results when applied to poorly trained
or otherwise deficient models.

Note that Figure \ref{fig:GPT-snorm-hist} also shows a couple anomalously  large Spectral Norms.
From Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
These layers appear to have a normalization, and therefore a differenc scale.
For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we layer all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier).
Here, we do not include them in our computed average metrics in Table \ref{table:nlp},
and do not include them in the histogram plot in Figure \ref{fig:GPT-snorm-hist}.

\paragraph{The Correlation Flow in GPT and GPT2} also differs significantly between GPT and GPT2.
Figure \ref{fig:gpt-alpha-layer} plots $\alpha$ vs the depth (i.e. a layer id) for each model.

@@

    \subfigure[Power Law Exponent $\alpha$]{
        \includegraphics[width=4.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \qquad
    \subfigure[log Spectral Norm $\log\Vert\mathbf{W}\Vert_{\infty}$]{
        \includegraphics[width=4.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \qquad
    \subfigure[log Alpha-Norm $\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$]{
        \includegraphics[width=4.5cm]{img/GPT-pnorm-depth.png}
        \label{fig:gpt-pnorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}
